Data exploration, quality control and statistical analysis of ChIP-exo experiments

### 2015-12-01 rmarkdown/slides.Rnw
finished slides for prelim

\section{ChIP-exo Data Exploration and Qualiy Control}

@@

  \begin{itemize}
\only<1>{
  \item Number of reads. Self-explanatory, the higher the better
}
  \item<2-> Number of reads.
\only<2>{
\item PCR bottleneck Coeff. Ratio of number of pos. to which
  {\color{RoyalBlue}EXACTLY} one reads maps and number of pos. to
  which {\color{RoyalBlue}AT LEAST} one reads maps }
\only<2>{
\begin{figure}[H]
  \centering
  \includegraphics[width = .75\textwidth]{../figs/for_paper/PBC_example.pdf}
\end{figure}
}
\item<3-> PCR bottleneck Coeff.
\only<3,4>{
\item Strand Cross-Correlation. 
\begin{align}
  y(\delta) = \sum_c w_c r\left[ n_c^+ \left(x + \frac{\delta}{2}
    \right), n_c^- \left( x- \frac{\delta}{2} \right)\right]
\nonumber
\end{align}
}
\only<3>{
  \begin{figure}[H]
    \centering
    \includegraphics[width = .4\textwidth]{../figs/for_paper/Kharchenko.png}
\caption{SCC explanation. Kharchenko et al., 2008 \cite{strandcc}}
  \end{figure}
}
\only<4>{
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth]{../figs/for_paper/SCC.png}
\caption{SCC as QC. Landt et al., 2012 \cite{encode_qc}}
\end{figure}
}
\item<5-> Strand Cross-Correlation.
\only<5>{
\item Normalized Strand Cross-Correlation. Ratio between the SCC when
  the shift is the fragment length and min. value of the SCC.
}
\end{itemize}
\end{frame}

@@

{\small
\begin{itemize}
\item A shows that high density regions are similar between ChIP-Seq
  and ChIP-exo but background regions are not.
\item The peak-pair assumption does hold in ChIP-exo data, but not
  locally since some regions show strand-imbalance.
\end{itemize}
}

@@

{\small
\begin{itemize}
\item There is a \emph{``phantom peak''} at read length.
\item In ChIP-Seq SCC is maximized at the unobserved fragment length.
\item In ChIP-exo, the \emph{``phantom peak''} and the fragment length
  summit are confounded.
\end{itemize}
}
\end{frame}

@@

 % \begin{itemize}
  % \item $\mbox{local-NSC}$
  % \item $\mbox{FSR} = \frac{\text{Nr. of fwd. strand reads in region}}{\text{Nr. of reads in region}}$
  % \end{itemize}

\begin{frame}
  \frametitle{ARC vs. URCR exploration}
\only<1>{
  \begin{itemize}
  \item $\mbox{ARC} = \frac{\text{Nr. of reads in the
        region}}{\text{Width of the region}}$
  \item $\mbox{URCR} = \frac{\text{Nr. of reads mapped to only one
        position in the region}}{\text{Nr. of reads in the region}}$
  \end{itemize}
}
\only<2>{
All regions
  \begin{figure}[H]
    \centering
    \includegraphics[width = .7\textwidth,page = 1]{../figs/for_paper/ARC_vs_URCR_example_npos.pdf}
  \end{figure}
}

\only<3>{
Mapped to more than 10 positions
  \begin{figure}[H]
    \centering
    \includegraphics[width = .7\textwidth,page = 2]{../figs/for_paper/ARC_vs_URCR_example_npos.pdf}
  \end{figure}
}

\only<4>{
Mapped to more than 30 positions
  \begin{figure}[H]
    \centering
    \includegraphics[width = .7\textwidth,page = 3]{../figs/for_paper/ARC_vs_URCR_example_npos.pdf}
  \end{figure}
}
\only<5>{
Mapped to more than 50 positions
  \begin{figure}[H]
    \centering
    \includegraphics[width = .7\textwidth,page = 4]{../figs/for_paper/ARC_vs_URCR_example_npos.pdf}
  \end{figure}
}

@@

% \only<9-10>{
%   \begin{figure}[H]
%     \centering
%     \includegraphics[width = .7\textwidth,page = 5]{../figs/for_paper/enrichment_example.pdf}
%   \end{figure}
% }

% \only<10>{
% \Put(120,200){\includegraphics[width = .6\textwidth,page = 5]{../figs/for_paper/coverage_example.pdf}}
% }

@@

\begin{frame}
  \frametitle{ARC vs. URCR exploration by replicate}
\only<1>{
All regions
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 1]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
\end{figure}
}

\only<2>{
Regions formed by reads aligned to more than 10 unique positions
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 3]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
\end{figure}
}

\only<3>{
Regions formed by reads aligned to more than 30 unique positions
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 4]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
\end{figure}
}

@@

\only<1>{
Calculate coverage for all regions
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 1]{../figs/for_paper/local_SCC_separated.pdf}
\end{figure}
}

\only<2>{
Calculate $\mbox{local-SCC}$ for regions
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 2]{../figs/for_paper/local_SCC_separated.pdf}
\end{figure}
}

\only<3>{
Fit local polynomial regression for the $\mbox{local-SCC}$ 
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 3]{../figs/for_paper/local_SCC_separated.pdf}
\end{figure}
}

@@

\begin{itemize}
\item $x_\delta$ is the shift.
\item $y_\delta$ is the local - SCC at shift $x_\delta$
\item $\epsilon \sim N(0, \sigma^2)$ i.i.d 
\end{itemize}

@@

  \begin{itemize}
  \item {\color{RoyalBlue} high:} Number of unique positions $ > 100$
  \item {\color{RoyalBlue} med:} Number of unique positions in $(50,100)$
  \item {\color{RoyalBlue} low:} Number of unique positions in $(20,50)$
  \end{itemize}
}
\end{frame}

@@

{\scriptsize
\begin{itemize}
\item 
\end{itemize}
}

@@

\begin{frame}[plain]
  \frametitle{Data structure}
Data available for $\sig$:
  \begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{../figs/for_paper/chip_explanation.pdf}
  \end{figure}

@@

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = .7\textwidth]{../figs/for_paper/algorithm_resolution.pdf}
\end{figure}

@@

  \begin{itemize}
  \item Our pipeline is capable of assessing the balance between
    sample enrichment and library complexity.
  \item We shown that the ``peak-pair'' assumption doesn't hold well
    in practice, and implemented a visualization capable of detecting
    strand imbalance.
  \item We updated dPeak, which makes a striking balance in
    sensitivity, specificity and spatial resolution.
  \item ChIP-exo and PET ChIP-Seq are comparable in resolution and
    sensitivity, and both outperform SET ChIP-Seq.
  \item We showed that with a fixed number of reads, ChIP-exo
    outperforms PET and SET ChIP-Seq.
  \item dPeak outperforms other algorithms in resolution.
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Future work}

\begin{itemize}
\item In the paper, we showed that there is a relationship between
  ChIP-exo tag counts and both mappability and GC content scores. We
  want to add a QC measure to the pipeline based on them.
\item We want to assess if ChIP-Nexus library complexity is actually
  higher than ChIP-exo's by using the $\mbox{local-NSC}$.
\item We have been studying E. Coli's transcription initiation
  complexes with PET ChIP-Seq. We want to improve this analysis by
  using ChIP-exo data.
\item Find a optimal strategy for labeling enhancer out of a
  predetermined list of regions in the genome by the use of active
  learning techniques.
\end{itemize}

@@

>>>>

\section{ChIP-exo Data Exploration and Quality Control}

@@

  \begin{itemize}
\only<1>{
\item Number of reads. Self-explanatory, the higher the better
  (usually).}
  \item<2-> Number of reads.
\only<2>{
\item PCR bottleneck Coeff. Ratio of number of pos. to which
  {\color{RoyalBlue}EXACTLY} one reads maps and number of pos. to
  which {\color{RoyalBlue}AT LEAST} one reads maps.}
\only<2>{
\begin{figure}[H]
  \centering
  \includegraphics[width = .75\textwidth]{../figs/for_paper/PBC_example.pdf}
\end{figure}
}
\item<3-> PCR bottleneck Coeff.
\only<3,4>{
\item Strand Cross-Correlation. 
\begin{align}
  y(\delta) = \sum_c w_c r\left[ n_c^+ \left(x + \frac{\delta}{2}
    \right), n_c^- \left( x- \frac{\delta}{2} \right)\right]
\nonumber
\end{align}
}
\only<3>{
  \begin{figure}[H]
    \centering
    \includegraphics[width = .4\textwidth]{../figs/for_paper/Kharchenko.png}
\caption{SCC explanation. Kharchenko et al., 2008 \cite{strandcc}}
  \end{figure}
}
\only<4>{
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth]{../figs/for_paper/SCC.png}
\caption{SCC as QC. Landt et al., 2012 \cite{encode_qc}}
\end{figure}
}
\item<5-> Strand Cross-Correlation.
\only<5>{
\item Normalized Strand Cross-Correlation. Ratio between the SCC when
  the shift is the fragment length and min. value of the SCC.
}
\item<6> Normalized Strand Cross-Correlation. 
\only<6>{
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth]{../figs/for_paper/SCC.png}
\caption{SCC as QC. Landt et al., 2012 \cite{encode_qc}}
\end{figure}
}
\end{itemize}
\end{frame}

@@

{\small
\begin{itemize}
\item A) shows that high density regions are similar between ChIP-Seq
  and ChIP-exo but background regions are not.
\item The peak-pair assumption does hold in ChIP-exo data, but not
  locally since some regions show strand-imbalance.
\end{itemize}
}

@@

{\small
\begin{itemize}
\item There is a \emph{``phantom peak''} at read length for ChIP-Seq (SE).
\item In ChIP-Seq, SCC is maximized at the unobserved fragment length.
\item In ChIP-exo, the \emph{``phantom peak''} and the fragment length
  summit are confounded.
\end{itemize}
}

@@

@@

@@

\begin{frame}
  \frametitle{ARC vs. URCR exploration by replicate}
\only<1>{
All regions:
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 1]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
\end{figure}
}

\only<2>{
Regions formed by reads aligned to more than 10 unique positions:
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 3]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
\end{figure}
}

\only<3>{
Regions formed by reads aligned to more than 30 unique positions:
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 4]{../figs/Carroll_mice_for_paper/FoxA1_enrichment.pdf}
\end{figure}
}

@@

\only<1>{
Calculate coverage for all regions:
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 1]{../figs/for_paper/local_SCC_separated.pdf}
\end{figure}
}

\only<2>{
Calculate $\mbox{local-SCC}$ for regions:
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 2]{../figs/for_paper/local_SCC_separated.pdf}
\end{figure}
}

\only<3>{
Fit local polynomial regression for the $\mbox{local-SCC}$ :
\begin{figure}[H]
  \centering
  \includegraphics[width = \textwidth,page = 3]{../figs/for_paper/local_SCC_separated.pdf}
\end{figure}
}

@@

\begin{itemize}
\item $x_\delta$ is the shift.
\item $y_\delta$ is the local - SCC at shift $x_\delta$.
\item $\epsilon \sim N(0, \sigma^2)$ i.i.d .
\end{itemize}

@@

  \begin{itemize}
  \item {\color{RoyalBlue} high:} Number of unique positions $ > 100$.
  \item {\color{RoyalBlue} med:} Number of unique positions in $(50,100)$.
  \item {\color{RoyalBlue} low:} Number of unique positions in $(20,50)$.
  \end{itemize}
}
\end{frame}

@@

{\scriptsize
\begin{itemize}
\only<1>{
\item The quantiles of the first replicate, approximate the median
  faster than the other replicate's quantiles.
 }
\only<2>{
\item B) shows that the majority of the low depth islands are formed
  exclusively by one strand.
}

\end{itemize}
}

@@

\begin{frame}[plain]
  \frametitle{Data structure}
Available for $\sig$ (courtesy of Prof. Landick's lab):
  \begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{../figs/for_paper/chip_explanation.pdf}
  \end{figure}

@@

\only<1>{
{\small
  \begin{itemize}
  \item There are more candidate regions for both ChIP-Seq samples,
    hence there are also more predicted events.
  \end{itemize}
}}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width = .6\textwidth]{../figs/for_paper/algorithm_resolution.pdf}
\end{figure}

{\scriptsize
  \begin{itemize}
  \item Resolution is defined as the min. absolute distance from the
    regulonDB annotation to an estimated binding site.
  \end{itemize}
}

@@

  \begin{itemize}
  \item Our pipeline is capable of assessing the balance between
    sample enrichment and library complexity.
  \item We shown that the ``peak-pair'' assumption does not hold well
    in practice, and implemented a visualization capable of detecting
    strand imbalance.
  \item We updated dPeak, which makes a striking balance in
    sensitivity, specificity and spatial resolution.
  \item ChIP-exo and PET ChIP-Seq are comparable in resolution and
    sensitivity, and both outperform SET ChIP-Seq.
  \item We showed that with a fixed number of reads, ChIP-exo
    outperforms PET and SET ChIP-Seq.
  \item dPeak outperforms other algorithms in resolution.
  \end{itemize}
\end{frame}

\begin{frame}[label=future]
\frametitle{Future work}

\begin{itemize}
\item In the paper, we showed that there is a relationship between
  ChIP-exo tag counts and both mappability and GC content scores. We
  want to add a QC measure to the pipeline based on
  them. \hyperlink{gcmap}{\beamerbutton{GC \& Mapp}}
\item We want to assess if ChIP-Nexus library complexity is actually
  higher than ChIP-exo's by using the
  $\mbox{local-NSC}$. \hyperlink{nexus}{\beamerbutton{ChIP-Nexus}}

\item We have been studying E. Coli's transcription initiation
  complexes with PET ChIP-Seq. We want to improve this analysis by
  using ChIP-exo data.  \hyperlink{tic}{\beamerbutton{TIC}}
\item Find a optimal strategy for labeling enhancer out of a
  predetermined list of regions in the genome by the use of active
  learning techniques.
\end{itemize}

@@

\begin{frame}[label=gcmap]
\frametitle{GC content and Mappability score bias}

\begin{figure}[H]
  \centering
  \includegraphics[width = .46\textwidth,page = 1]{../figs/for_paper/eukaryotic_bias_CTCF.pdf}
  \includegraphics[width = .46\textwidth,page = 2]{../figs/for_paper/eukaryotic_bias_CTCF.pdf}

  \begin{itemize}
  \item ChIP tag counts increase linearly as mappability scores
    increases.
  \item ChIP tag counts increase linearly as GC content score
    increases when GC content is less than 0.6 and then ChIP tag
    counts decrease as GC content
    increases. \hyperlink{future}{\beamerbutton{back}}

  \end{itemize}

\end{figure}

\begin{frame}[plain,label=nexus]
\frametitle{ChIP-exo and ChIP- Nexus}

\begin{columns}
\column{.5\textwidth}
  \begin{figure}[H]
    \includegraphics[width = .9\textwidth]{../vignettes/exo-protocol.png}
\caption{ChIP-exo explanation, Mendenhall and Bernstein, 2012 \cite{exo_gb}}
  \end{figure}
\column{.5\textwidth}

\begin{itemize}
\item ChIP-Nexus uses a random barcode for adapter $\#$ 1.
\item Before step 5 uses self-circularization to attach the random
  barcode to the 3' end, and uses it instead of adapter $\#$ 2.

\end{itemize}
  \begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{../figs/for_paper/nexus.png}
    \caption{ChIP-Nexus example, He et al., 2015, \cite{chipnexus}
      \hyperlink{future}{\beamerbutton{back}}}
  \end{figure}

\end{columns}

\begin{frame}[label=tic]
\frametitle{Transcription Initiation Complexes}

  \begin{figure}[H]
  \centering
 \includegraphics[width = \textwidth]{/p/keles/ChIPexo/volume5/OpenClosedComplexes/figs/filtered/heatmaps_centroid.png}
 \caption{Centered and normalized heatmaps for PE ChIP-Seq peaks of
   $\sig$ and $\beta'_f$ under rif0 and rif20. The regions are
   organized by the centroids clustering of the $\beta'_f$-rif0
   profiles. \hyperlink{future}{\beamerbutton{back}}}
  \label{fig:hm1}
\end{figure}
